---
title: "IBM Attrition Data Set"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/k.vincent/Desktop/predictive modeling/Project")
attribution = read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
```

####__Group Project  - July 23, 2018__


This is a course project in MS Business Analytics, seeking to determine which factors and to what degree each factor is driving employee attrition.
We get the data from IBM Data Science Team Kaggle.com - HR Analytics

Setups: 
```{r}
library(ggplot2)
library(caret)
library(MASS)
library(tree)
library(knitr)
library(randomForest)
library(gbm)
library(corrplot)
library(glmnet)
library(e1071)
```


<br>

####__1. Data Cleaning__

Overview
```{r}
# Overview: 1470 obs. of  35 variables
str(attribution)
```

Change Strings to multi-level factors:
```{r}
attribution$Attrition = as.factor(attribution$Attrition)
attribution$BusinessTravel = as.factor(attribution$BusinessTravel)
attribution$Department = as.factor(attribution$Department)
attribution$EducationField = as.factor(attribution$EducationField)
attribution$Gender = as.factor(attribution$Gender)
attribution$JobRole = as.factor(attribution$JobRole)
attribution$MaritalStatus = as.factor(attribution$MaritalStatus)
attribution$OverTime = as.factor(attribution$OverTime)
```

```{r}
#na?
any(is.na(attribution))
```

Remove Columns: Over18, EmployeeCount, EmployeeNumber, standardhour, which are not relevant to Attrition. 
```{r}
#na?
attribution = attribution[ , c(-9,-10,-22, -27)]
```


####__2. EDA: Correlation__

```{r}
numerical = unlist(lapply(attribution, is.numeric))
M = cor(attribution[, numerical])
corrplot.mixed(M, tl.cex=0.6)
```

As a result:<br>
JobLevel & MonthlyIncome: 0.95 <br>
JobLevel & WorkingYears: 0.78<br>
PercentSalaryHike & Performance Rating: 0.77<br>
MonthlyIncome & Working Years: 0.77<br>
YearsatCompany & YearsCurrManager: 0.77<br>
YearsatCompany & YearInCurrentRole: 0.76<br>
YearInCurrentRole & YearsCurrManager: 0.71<br>
Age & WorkingYears: 0.68<br>
WorkingYears & YearsatCompany: 0.63<br>
YearsatCompany & YearssinceLastPromotion: 0.62<br>

####__3. EDA: Visualization of each independent variable against Attrition__
```{r}
ggplot(attribution, aes(x=Age, fill=Attrition)) + geom_density(alpha=0.3)

# Those who travel frequently are more likely to leave
g = ggplot(attribution, aes(x = Attrition, group = BusinessTravel)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g +facet_grid(.~BusinessTravel) + ggtitle("BusinessTravel")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# attrition under dailyRate: people with lower dailyrate tend to leave
ggplot(attribution, aes(x=DailyRate, fill=Attrition)) + geom_density(alpha=0.3)

# People in Sales are more likely to leave
g = ggplot(attribution, aes(x =Attrition , group = Department))+geom_bar(aes(y = ..prop..), stat="count", fill = "#FFA500", alpha=0.3)
g +facet_grid(.~Department) + ggtitle("Department")+theme_bw() + geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# People live far way tend to leave
ggplot(attribution, aes(x=DistanceFromHome, fill=Attrition)) + geom_density(alpha=0.3)

# Higher the education level,more likely to stay
g = ggplot(attribution, aes(x = Attrition, group = Education)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~Education) + ggtitle("Education")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Those with HR, Marketing and Technical education background are more likely to leave 
g = ggplot(attribution, aes(x = Attrition, group = EducationField)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~EducationField) + ggtitle("EducationField")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Environment satisfaction is important
g = ggplot(attribution, aes(x = Attrition, group = EnvironmentSatisfaction)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~EnvironmentSatisfaction) + ggtitle("EnvironmentSatisfaction")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Males are more likely to leave
g = ggplot(attribution, aes(x = Attrition, group = Gender)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~Gender) + ggtitle("Gender")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Hourly Rate: lower rate made people leave
ggplot(attribution, aes(x=HourlyRate, fill=Attrition)) + geom_density(alpha=0.3)

# Job involvement matters(a lot)
g = ggplot(attribution, aes(x = Attrition, group = JobInvolvement)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~JobInvolvement) + ggtitle("JobInvolvement")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Job Level matters
g = ggplot(attribution, aes(x = Attrition, group = JobLevel)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~JobLevel) + ggtitle("JobLevel")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Job role matters(HR, laboratory technician and sales representatives are more likely to leave)
g = ggplot(attribution, aes(x = Attrition, group = JobRole)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~JobRole) + ggtitle("JobRole")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", hjust=0.3,vjust = -.3)

# Job satisfaction matters(a lot)
g = ggplot(attribution, aes(x = Attrition, group = JobSatisfaction)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~JobSatisfaction) + ggtitle("JobSatisfaction")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Single people tend to leave
g = ggplot(attribution, aes(x = Attrition, group = MaritalStatus)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~MaritalStatus) + ggtitle("MaritalStatus")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Monthly income Yes
ggplot(attribution, aes(x=MonthlyIncome, fill=Attrition)) + geom_density(alpha=0.3)

# Monthly Rate : not clear
ggplot(attribution, aes(x=MonthlyRate, fill=Attrition)) + geom_density(alpha=0.3)

# Num of company worked: Yes
ggplot(attribution, aes(x=NumCompaniesWorked, fill=Attrition)) + geom_density(alpha=0.3)

# Those who work overtime are more likely to leave
g = ggplot(attribution, aes(x =Attrition , group = OverTime)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~OverTime) + ggtitle("OverTime")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Percent Salary Hike: yes
ggplot(attribution, aes(x=PercentSalaryHike, fill=Attrition)) + geom_density(alpha=0.3)

# Performance Rating： no
ggplot(attribution, aes(x=PerformanceRating, fill=Attrition)) + geom_density(alpha=0.3)


ggplot(attribution, aes(x=StockOptionLevel, fill=Attrition)) + geom_density(alpha=0.3)


ggplot(attribution, aes(x=TotalWorkingYears, fill=Attrition)) + geom_density(alpha=0.3)

# TrainingTimesLastYear: no
ggplot(attribution, aes(x=TrainingTimesLastYear, fill=Attrition)) + geom_density(alpha=0.3)

# Worklifebalance(higher worklifebalance score-->stay, but the highest score4 is an exception)
g = ggplot(attribution, aes(x = Attrition, group = WorkLifeBalance)) + geom_bar(aes(y = ..prop..), stat="count", fill="#FFA500", alpha=0.3)
g + facet_grid(.~WorkLifeBalance) + ggtitle("WorkLifeBalance")+theme_bw()+geom_text(aes(label = scales::percent(..prop..), y = ..prop.. ), stat= "count", vjust = -.3)

# Years at Company matters
ggplot(attribution, aes(x=YearsAtCompany, fill=Attrition)) + geom_density(alpha=0.3)

ggplot(attribution, aes(x=YearsInCurrentRole, fill=Attrition)) + geom_density(alpha=0.3)

# YearSinceLastPromotion: no
ggplot(attribution, aes(x=YearsSinceLastPromotion, fill=Attrition)) + geom_density(alpha=0.3)

# YearswithCurrent manager matters
ggplot(attribution, aes(x=YearsWithCurrManager, fill=Attrition)) + geom_density(alpha=0.3)
```

####__4. Modeling__

We can consider logistic regression, boosting, random forest, knn.'

As the data nature is highly unbalanced: most of them have attrition "No", we decided to add a threshold for the models.
For each model, we tested different threshold and pick the one that makes the best prediction.

We use accuracy, percentage of people who leaves / people we predict to leave, and percentage of people leave / people we predict to stay to compare the models because our emphasis on predicting people who intend to leave the company.

######__Performance Measurement Criteria__
```{R}
calc_acc = function(actual, predicted) {
mean(actual == predicted)
}
# how many people actually leaves when we predicted them to leave
calc_stay = function(TP, FP){
TP / (TP + FP)
}
calc_leave = function(y, x){
x / (x + y)
}
```

We use 70% of our dataset as traning data.
```{R}
set.seed(432)
index = sample(nrow(attribution), size = trunc(0.7 * nrow(attribution)))
32
train_data= attribution[index, ]
test_data = attribution[-index, ]
```

######__Logistic__
First, We start with Logistic Regression and we used mixed selection to get 18 important variables. From the model summary, we can see that each variable is significant.

Modeling and Feature Selection using step-wise: 
```{R}
set.seed(432)
null = glm(Attrition~1, data=train_data, family="binomial")
log.fit =glm(Attrition~., data=train_data, family="binomial")
regboth = step(null, scope=formula(log.fit), direction="both", trace=0)
log.fit = glm(Attrition ~ OverTime + JobRole + JobInvolvement +
MaritalStatus + JobSatisfaction + EnvironmentSatisfaction +
BusinessTravel + DistanceFromHome + YearsInCurrentRole +
YearsSinceLastPromotion + TrainingTimesLastYear + Age +
NumCompaniesWorked + RelationshipSatisfaction + WorkLifeBalance +
YearsWithCurrManager + YearsAtCompany + TotalWorkingYears, data = train_data, family="binomial")
summary(log.fit)
```

After testing, a threshold of 0.7 maximize out-of-sample accuracy. 
Threshold setting: 
```{R}
log_pred = ifelse(predict(log.fit, newdata = test_data, type = "response") >= 0.7, 'Yes', 'No')
t1 = table(predicted = log_pred, actual = test_data$Attrition)
t1

```

######__KNN__
In Knn, we use the features above and apply 10-fold validation to tune the parameter k of knn.
```{R}
library(class) 
library(kknn) 

df_non_dummy<- attribution[,c(-3,-5,-8,-10,-14,-16,-20)]
x<- scale(df_non_dummy[,-2])
dfready<- data.frame(x)
dfready[,24]<- df_non_dummy[,2]
colnames(dfready)[24]<- "Attrition"
train <- sample(1:1470,0.8*1470)
a = 1:1470
test = a[-train]
nearest5 <- kknn(Attrition~., dfready[train,],dfready[-train,], k=5)


fit<- function(kknnprob, thres = 0.5){
  ft<- factor(c("Yes","No"))
  for(i in 1:nrow(kknnprob)){
    if(kknnprob[,2][i]>thres){
      ft[i] = "Yes"
    }
    else{
      ft[i]= "No"
    }
  }
  return(ft)
}

dfready<-data.frame(c(dfready, attribution[,c(3,5,8,10,14,16,20)]))
#model 
"Attrition ~ Age + OverTime + MaritalStatus + JobRole + BusinessTravel + 
    NumCompaniesWorked + JobInvolvement + JobSatisfaction + DistanceFromHome + 
YearsInCurrentRole + YearsSinceLastPromotion + TotalWorkingYears + 
RelationshipSatisfaction + DailyRate + TrainingTimesLastYear + 
StockOptionLevel"

n = nrow(dfready)
kcv = 10
n0 = round(n/kcv,0)

out_MSE = matrix(0,kcv,100)

used = NULL
set = 1:n

for(j in 1:kcv){
  
  if(n0<length(set)){val = sample(set,n0)}
  
  if(n0>=length(set)){val=set}
  
  train_i = dfready[-val,]
  test_i = dfready[val,]
  
  
  for(i in 11:110){
    
    near = kknn(Attrition ~ Age
                +                     NumCompaniesWorked + JobInvolvement + JobSatisfaction + DistanceFromHome + 
                  +                     YearsInCurrentRole + YearsSinceLastPromotion + TotalWorkingYears + 
                  +                     RelationshipSatisfaction + DailyRate + TrainingTimesLastYear + 
                  +                     StockOptionLevel + BusinessTravel + Department + EducationField + Gender+JobRole + MaritalStatus + OverTime ,train_i,test_i,k=i,kernel = "rectangular")
    localCount = 0
    ft<- fit(near$prob, 0.27)
    for(g in 1:nrow(test_i)){
      if(test_i[g,"Attrition"]!=ft[g]){
        localCount = localCount+1
      }
    }
    
    out_MSE[j,i-10] = localCount
  }
  
  used = union(used,val)
  set = (1:n)[-used]
  
  cat(j,'\n')
  
}
mMSE = apply(out_MSE,2,mean)
best = which.min(mMSE)
print(1-(mMSE[best]/nrow(test_i)))


plot(log(1/(1:100)),sqrt(mMSE),xlab="Complexity (log(1/k))",ylab="out-of-sample RMSE",col=4,lwd=2,type="l",cex.lab=1.2,main=paste("kfold(",kcv,")"))
best = which.min(mMSE)
text(log(1/best),sqrt(mMSE[best])-0.005,paste("k=",best),col=2,cex=1)
text(log(1/11)+2.25,sqrt(mMSE[2])-0.3+0.27,paste("k=",11),col=2,cex=1)
text(log(1/110)+0.3,sqrt(mMSE[100]),paste("k=",110),col=2,cex=1)
```

######__Boosting__
```{R}
set.seed(432)
train_data_copy = train_data
train_data_copy$Attrition = ifelse(train_data_copy$Attrition == "No", 0, 1)
boosting_1 = gbm(Attrition~., data=train_data_copy, distribution="bernoulli", n.trees=1000, shrinkage = 0.01)
summary(boosting_1)
```

After adjusting the shrinkage and n.minobsinnode:
```{R}
set.seed(432)
boosting_1 = gbm(Attrition~.-PerformanceRating-Gender-Department, data=train_data_copy, distribution="bernoulli", n.trees=1000, shrinkage =0.01, n.minobsinnode =4)
boo_pred = ifelse(predict(boosting_1, newdata = test_data, n.trees = 1000, type="response")>0.45, 'Yes', 'No')
t6=table(predicted = boo_pred, actual = test_data$Attrition)
t6
```

######__Random Forest__
We start by fitting all the variables in the model and then take out the least important variable according to importance plots, until all existing features have positive importance. 
```{R}
set.seed(432)
rf_1 = randomForest(Attrition~., data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_1, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_1)

set.seed(432)
rf_2 = randomForest(Attrition~.-MonthlyRate, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_2, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_2)

set.seed(432)
rf_3 = randomForest(Attrition~.-MonthlyRate - PerformanceRating, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_3, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_3)

set.seed(432)
rf_4 = randomForest(Attrition~.-MonthlyRate - PerformanceRating - TrainingTimesLastYear, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_4, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_4)

set.seed(432)
rf_5 = randomForest(Attrition~.-MonthlyRate - PerformanceRating - TrainingTimesLastYear - Gender, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_5, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_5)

set.seed(432)
rf_6 = randomForest(Attrition~.-MonthlyRate - PerformanceRating - TrainingTimesLastYear - Gender - WorkLifeBalance , data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_6, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_6)

set.seed(432)
rf_7 = randomForest(Attrition~.-MonthlyRate - PerformanceRating - TrainingTimesLastYear - Gender - WorkLifeBalance - PercentSalaryHike, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_7, newdata = test_data, type = "prob")[ ,1] >= 0.7, 'No', 'Yes')
table(rf_pred, test_data$Attrition)
varImpPlot(rf_7)

set.seed(432)
rf_8 = randomForest(Attrition~.-MonthlyRate - PerformanceRating - TrainingTimesLastYear - Gender - WorkLifeBalance - PercentSalaryHike - DailyRate, data = train_data, importance=TRUE)
rf_pred = ifelse(predict(rf_8, newdata = test_data, type = "prob")[ ,1] >= 0.6, 'No', 'Yes')
t4=table(rf_pred, test_data$Attrition)
t4
varImpPlot(rf_8)
```


######__naivebayes__

```{R}
set.seed(56)
nb_1 = naiveBayes(Attrition~., data = train_data, prior=c(1233, 237)/1470)
nb_pred = ifelse(predict(nb_1, test_data, type = "raw")[ ,1]>=0.1, 'No', 'Yes')
calc_acc(nb_pred, test_data$Attrition)

t7=table(predicted = nb_pred, actual = test_data$Attrition)
t7
```


To Summary, 5 models were considered and tuned with feature selection and/or parameter validation. For model summary, review model summary file. 







